---
output: html_document
---
Tornados Have the Most Impact Among Severe Weather Events on Public Health and Economy in the United States
========================================================
#### Andrey Alferov
#### Sunday, March 22, 2015

Synopsis
========================================================
In this report we aim to answer the following questions:

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

To achieve it, we use the NOAA Storm Database. From this data, we found out that the most harmful type of events to population health and economy are **Tornados**. **Excessive Heat** followed by **Flash Flood**, **Heat** and **Lightning** are the next most harmful events with respect to population health. **Flash Flood**, **Tstm Wind**, **Flood** and **Thunderstorm Wind** caused the most ecomonomy damages after **Tornados**.

Data Processing
========================================================

### Load required libraries
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
```

### Download dataset file, unzip it and load in dataframe
```{r, warning=FALSE, message=FALSE, cache = TRUE}
setwd("D:/Coursera/Data Science/Reproducible Research/Perr assessment 2")

if(!file.exists("./data"))
{dir.create("./data")}

url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- "StormData.csv.bz2"

download.file(url, destfile = fileName)

stormData <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = FALSE)
```

### Convert to dplyr table
```{r}
stormData <- tbl_df(stormData)
```

### Select only fields required for analysis: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
```{r}
stormData %>% 
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) -> data
```

## Check EVTYPE dimension for cleanliness and clean it if neccessary
### Check EVTYPE column for cleanliness
```{r}
unique(data$EVTYPE)
```
As we can see there are dirty values containing numbers and odd strings like 'Summary'

### Perform some cleaning of EVTYPE column values to make the analysis more reliable
#### Convert strings to upper register, remove leading and trailing spaces, odd strings 'Summary' and numbers
```{r}
data$EVTYPE <- toupper(data$EVTYPE)
data$EVTYPE <- str_trim(data$EVTYPE, side = "both")
data <- data[-grep("SUMMARY", data$EVTYPE), ]
data$EVTYPE <- gsub("[0-9]+", "", data$EVTYPE)
```
Now we have much more consistent values in EVTYPE dimension
```{r}
unique(data$EVTYPE)
```

## Perform analisys of which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
### Check FATALITIES and INJURIES measures for NA values
```{r}
summarize(data, isna_fatalities = sum(is.na(FATALITIES) == TRUE), isna_injuries = sum(is.na(INJURIES) == TRUE))
```
There are no NA values in these measures, thus we don't have to deal with missing values.

### Summarise fatalities and injuries by event type
```{r}
data %>%
  group_by(EVTYPE) %>%
    summarise(fatalities_sum = sum(FATALITIES), injuries_sum = sum(INJURIES)) -> fatalities_by_event_type
```

### Find event types caused the top 5 of fatalities & injuries
```{r}
fatalities_by_event_type %>%
  arrange(desc(fatalities_sum)) %>%
    head(5) -> top5_fatalities_by_event_type

top5_fatalities_by_event_type

fatalities_by_event_type %>%
  arrange(desc(injuries_sum)) %>%
    head(5) -> top5_injuries_by_event_type

top5_injuries_by_event_type
```

## Perform analisys of which types of events (as indicated in the EVTYPE variable) which types of events 
## have the greatest economic consequences?

### Check PROPDMGEXP and CROPDMGEXP dimensions for cleanliness and clean them if neccessary
```{r}
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

#### Convert strings to upper register to make them uniform
```{r}
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)
```

#### Create function to convert character codes in PROPDMGEXP and CROPDMGEXP dimensions to numeric values of multypliers
```{r}
convert_to_number <- function(val) { 
  if (is.na(val))
    1
  else if (val == "H")
    100
  else if (val == "K")
    1000
  else if (val == "M")
    1000000
  else if (val == "B")
    1000000000
  else if (val == "?" | val == "+" | val == "-")
    1
  else {
    n = as.integer(val)
    if (!is.na(n))
      10**n
    else
      1
  }        
}
```

### Check PROPDMG and CROPDMG measures for NA values
```{r}
summarize(data, isna_propdmg = sum(is.na(PROPDMG) == TRUE), isna_cropdmg = sum(is.na(CROPDMG) == TRUE))
```
There are no NA values in these measures, thus we don't have to deal with missing values.

### Create new variables containing property and crop damages values multiplied by corresponding multiplicator
```{r, warning=FALSE, message=FALSE}
data %>%
  mutate(property_damages = PROPDMG * convert_to_number(PROPDMGEXP), 
         crop_damages = CROPDMG * convert_to_number(CROPDMGEXP) ) -> data
```

### Create new variable containing overall economic damages calculated as a sum of property and crop damages
```{r}
data %>%
  mutate(total_damages = property_damages + crop_damages) -> data
```

### Summarise total damages by event type
```{r}
data %>%
  group_by(EVTYPE) %>%
  summarise(total_damages_sum = sum(total_damages)) -> damages_by_event_type
```

### Find event types caused the top 5 of total damages
```{r}
damages_by_event_type %>%
  arrange(desc(total_damages_sum)) %>%
  head(5) -> top5_total_damages_by_event_type

top5_total_damages_by_event_type
```

Results
========================================================
## Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
### Draw a barplot displaying top 5 types of events caused fatalities across the U.S.
```{r}
p1 <- ggplot(top5_fatalities_by_event_type, aes(reorder(EVTYPE, -fatalities_sum), fatalities_sum, fill = EVTYPE)) + geom_bar(stat = "identity")
p1 <- p1 + ylab("Fatalities") + xlab("Event type") + ggtitle("Top 5 types of events caused fatalities across the U.S.")
p1 <- p1 + theme(legend.position="none")
p1
```

### Draw a barplot displaying top 5 types of events caused injuries across the U.S.
```{r}
p2 <- ggplot(top5_injuries_by_event_type, aes(reorder(EVTYPE, -injuries_sum), injuries_sum, fill = EVTYPE)) + geom_bar(stat = "identity")
p2 <- p2 + ylab("Injuries") + xlab("Event type") + ggtitle("Top 5 types of events caused injuries across the U.S.")
p2 <- p2 + theme(legend.position="none")
p2
```

As we can see on the barplots above **Tordanos** are most harmful with respect to population health both in terms of fatalities and injuries. 
**Excessive Heat**, **Flash Flood**, **Heat** and **Lightning** are the next most harmful events in terms of fatalities.
**Tstm Wind**, **Flood**, **Excessive Heat** and **Lightning** are the next most harmful events in terms of injuries.

## Across the United States, which types of events have the greatest economic consequences?
### Draw a barplot displaying top 5 types of events caused economic damages across the U.S.
```{r}
p3 <- ggplot(top5_total_damages_by_event_type, aes(reorder(EVTYPE, -total_damages_sum), total_damages_sum / 1000000, fill = EVTYPE)) + geom_bar(stat = "identity")
p3 <- p3 + ylab("Economic damages (millions dollars)") + xlab("Event type") + ggtitle("Top 5 types of events caused economic damages across the U.S.")
p3 <- p3 + theme(legend.position="none")
p3
```

As we can see on the barplot above **Tordanos** have the greatest economic consequences.
**Flash Flood**, **Tstm Wind**, **Flood** and **Thunderstorm Wind** have the greatest economic consequences after **Tornados**.
